CF Registry Analysis - ROC

### 2025-11-19 Christine Chan/CF Registry/cf_registry_ROC_analysis.qmd
piecewise models for HbA1c analysis

```{r setup}
#| include: false
library(tidyverse)
library(ggplot2)
library(gtsummary)
library(gt)
library(pROC)
library(ggrepel)
library(childsds)
library(labelled)
home_dir <- switch(
  Sys.info()["sysname"],
  "Darwin" = "/Users/tim/Library/CloudStorage/OneDrive-TheUniversityofColoradoDenver/Vigers/CF/Christine Chan/CF Registry",
  "Windows" = "C:/Users/tim/OneDrive - The University of Colorado Denver/Vigers/CF/Christine Chan/CF Registry",
  "Linux" = "/home/tim/OneDrive/Vigers/CF/Christine Chan/CF Registry"
)
github_dir <- switch(
  Sys.info()["sysname"],
  "Darwin" = "/Users/timvigers/GitHub",
  "Windows" = "C:/Users/Tim/GitHub",
  "Linux" = "/home/tim/GitHub"
)
knitr::opts_knit$set(root.dir = home_dir)
set.seed(1017)
```

@@

library(Hmisc)
library(lmerTest)

@@

  summarise(first_cfrd = min(ReviewYear))

@@

```{r}
#| warning: false
#| message: false
# Plot by diagnosis type
mod_df = cfrd_only %>%
  select(id, a1c_ogtt) %>%
  rename(eDWID = id)
mod_df = left_join(
  annual %>% mutate(eDWID = as.numeric(eDWID)),
  mod_df,
  by = join_by(eDWID)
)
mod_df = mod_df %>%
  group_by(eDWID) %>%
  mutate(years_to_cfrd = ReviewYear - first_cfrd)
var_label(mod_df$years_to_cfrd) = "Years to CFRD"
ggplot(
  mod_df %>% filter(!is.na(a1c_ogtt)),
  aes(
    y = A_FEV1_pct_predicted,
    x = years_to_cfrd,
    color = a1c_ogtt,
    group = a1c_ogtt
  )
) +
  xlim(c(-10, 10)) +
  geom_point(alpha = 0.1, shape = ".") +
  geom_smooth(formula = y ~ s(x, bs = "cr")) +
  theme_bw() +
  theme(legend.title = element_blank())
```

@@

```{r}
# Fit model
mod_df$x = mod_df$years_to_cfrd * (mod_df$years_to_cfrd >= 0)
label(mod_df$x) = "Change in Slope Post-CFRD"
label(mod_df$years_to_cfrd) = "Slope Pre-CFRD"
label(mod_df$a1c_ogtt) = "Diagnosis Method"
m <- lmer(
  A_FEV1_pct_predicted ~ years_to_cfrd *
    a1c_ogtt +
    x * a1c_ogtt +
    (1 | eDWID),
  data = mod_df
)
# Population-averaged plot

# Model table
m |> tbl_regression()
```

>>>>

```{r setup}
#| include: false
library(ggplot2)
library(gtsummary)
library(gt)
library(pROC)
library(ggrepel)
library(childsds)
library(labelled)
library(Hmisc)
library(lmerTest)
library(multcomp)
library(tidyverse)
home_dir <- switch(
  Sys.info()["sysname"],
  "Darwin" = "/Users/tim/Library/CloudStorage/OneDrive-TheUniversityofColoradoDenver/Vigers/CF/Christine Chan/CF Registry",
  "Windows" = "C:/Users/tim/OneDrive - The University of Colorado Denver/Vigers/CF/Christine Chan/CF Registry",
  "Linux" = "/home/tim/OneDrive/Vigers/CF/Christine Chan/CF Registry"
)
github_dir <- switch(
  Sys.info()["sysname"],
  "Darwin" = "/Users/timvigers/GitHub",
  "Windows" = "C:/Users/Tim/GitHub",
  "Linux" = "/home/tim/GitHub"
)
knitr::opts_knit$set(root.dir = home_dir)
set.seed(1017)
```

@@

@@

  summarise(first_cfrd = min(ReviewYear), age_dx = min(Age_YrEnd))

@@

```{r}
#| warning: false
#| message: false
# Plot by diagnosis type
mod_df = cfrd_only %>%
  select(id, a1c_ogtt, Age_YrEnd) %>%
  rename(eDWID = id)
mod_df = left_join(
  annual %>% select(-Age_YrEnd) |> mutate(eDWID = as.numeric(eDWID)),
  mod_df,
  by = join_by(eDWID)
)
mod_df = mod_df %>%
  drop_na(first_cfrd) |>
  group_by(eDWID) %>%
  mutate(years_to_cfrd = ReviewYear - first_cfrd)
var_label(mod_df$years_to_cfrd) = "Years to CFRD"
var_label(mod_df$Age_YrEnd) = "Age at Diagnosis"
ggplot(
  mod_df %>% filter(!is.na(a1c_ogtt)),
  aes(
    y = A_FEV1_pct_predicted,
    x = years_to_cfrd,
    color = a1c_ogtt,
    group = a1c_ogtt
  )
) +
  xlab("Years from CFRD") +
  xlim(c(-10, 10)) +
  geom_point(alpha = 0.1, shape = ".") +
  geom_smooth(formula = y ~ s(x, bs = "cr")) +
  theme_bw() +
  theme(legend.title = element_blank())
```

@@

There were not enough pancreatic sufficient people in this dataset to include that varaible in the model.

```{r}
# Fit model - adjust for age at dx, sex, genotype (severe vs mild), pancreatic
# insufficiency, age at CF diagnosis (do we know this?), number of clinic
# visits/year, use of CFTR modulators
mod_df$x = mod_df$years_to_cfrd * (mod_df$years_to_cfrd >= 0)
label(mod_df$x) = "Change in Slope Post-CFRD"
label(mod_df$years_to_cfrd) = "Slope Pre-CFRD"
label(mod_df$a1c_ogtt) = "Diagnosis Method"
label(mod_df$Age_YrEnd) = "Age at Diagnosis"
m <- lmer(
  A_FEV1_pct_predicted ~ years_to_cfrd *
    a1c_ogtt +
    x * a1c_ogtt +
    Age_YrEnd +
    Gender +
    MutClass +
    A_Vx770 +
    A_VX809comb +
    (1 | eDWID),
  data = mod_df
)
# Population-averaged plot
pop_dat <- data.frame(expand.grid(
  years_to_cfrd = -10:10,
  a1c_ogtt = unique(mod_df$a1c_ogtt),
  Age_YrEnd = mean(mod_df$Age_YrEnd, na.rm = T),
  Gender = "M",
  MutClass = "1-3",
  A_Vx770 = "No",
  A_VX809comb = "No"
))
pop_dat$x <- (pop_dat$years_to_cfrd >= 0) * pop_dat$years_to_cfrd
pred <- predict(m, re.form = NA, newdata = pop_dat, se.fit = T)
pop_dat$pred <- pred$fit
pop_dat$ll <- pop_dat$pred - (1.96 * pred$se.fit)
pop_dat$ul <- pop_dat$pred + (1.96 * pred$se.fit)
ggplot(
  pop_dat,
  aes(x = years_to_cfrd, y = pred, group = a1c_ogtt, colour = a1c_ogtt)
) +
  geom_point() +
  geom_errorbar(aes(ymin = ll, ymax = ul), width = 0.1) +
  geom_line() +
  xlab("Years from CFRD") +
  ylab("FEV1 % Predicted") +
  theme_classic(base_size = 15)
# Model table
m |> tbl_regression()
# Contrasts
neither_pre <- rep(0, nrow(summary(m)$coefficients))
neither_change <- rep(0, nrow(summary(m)$coefficients))
neither_pre[which(rownames(summary(m)$coefficients) == "years_to_cfrd")] = 1
neither_change[which(rownames(summary(m)$coefficients) == "x")] = 1
a1c_pre <- rep(0, nrow(summary(m)$coefficients))
a1c_change <- rep(0, nrow(summary(m)$coefficients))
a1c_pre[which(
  rownames(summary(m)$coefficients) %in%
    c("years_to_cfrd", "years_to_cfrd:a1c_ogttHbA1c Only")
)] = 1
a1c_change[which(
  rownames(summary(m)$coefficients) %in%
    c("x", "a1c_ogttHbA1c Only:x")
)] = 1
ogtt_pre <- rep(0, nrow(summary(m)$coefficients))
ogtt_change <- rep(0, nrow(summary(m)$coefficients))
ogtt_pre[which(
  rownames(summary(m)$coefficients) %in%
    c("years_to_cfrd", "years_to_cfrd:a1c_ogttOGTT")
)] = 1
ogtt_change[which(
  rownames(summary(m)$coefficients) %in%
    c("x", "a1c_ogttOGTT:x")
)] = 1
contr <- rbind(
  "HbA1c Only Pre-CFRD" = a1c_pre,
  "HbA1c Only Post-CFRD" = a1c_pre + a1c_change,
  "OGTT Pre-CFRD" = ogtt_pre,
  "OGTT Post-CFRD" = ogtt_pre + ogtt_change,
  "Neither Pre-CFRD" = neither_pre,
  "Neither Post-CFRD" = neither_pre + neither_change
)
change_mod_slopes <- glht(m, linfct = contr)
change_mod_slopes_ci <- confint(change_mod_slopes, calpha = univariate_calpha())
change_mod_slopes_ci$confint %>%
  as.data.frame() %>%
  gt(caption = "Slopes by Diagnosis Method", rownames_to_stub = T) %>%
  fmt_number(decimals = 2)
```

## FVC trends by diagnosis method

```{r} 
ggplot(
  mod_df %>% filter(!is.na(a1c_ogtt)),
  aes(
    y = A_FVC_pct_predicted,
    x = years_to_cfrd,
    color = a1c_ogtt,
    group = a1c_ogtt
  )
) +
  xlab("Years from CFRD") +
  xlim(c(-10, 10)) +
  geom_point(alpha = 0.1, shape = ".") +
  geom_smooth(formula = y ~ s(x, bs = "cr")) +
  theme_bw() +
  theme(legend.title = element_blank())
```

There were not enough pancreatic sufficient people in this dataset to include that varaible in the model.

```{r}
# Fit model - adjust for age at dx, sex, genotype (severe vs mild), pancreatic
# insufficiency, age at CF diagnosis (do we know this?), number of clinic
# visits/year, use of CFTR modulators
mod_df$x = mod_df$years_to_cfrd * (mod_df$years_to_cfrd >= 0)
label(mod_df$x) = "Change in Slope Post-CFRD"
label(mod_df$years_to_cfrd) = "Slope Pre-CFRD"
label(mod_df$a1c_ogtt) = "Diagnosis Method"
label(mod_df$Age_YrEnd) = "Age at Diagnosis"
m <- lmer(
  A_FVC_pct_predicted ~ years_to_cfrd *
    a1c_ogtt +
    x * a1c_ogtt +
    Age_YrEnd +
    Gender +
    MutClass +
    A_Vx770 +
    A_VX809comb +
    (1 | eDWID),
  data = mod_df
)
# Population-averaged plot
pop_dat <- data.frame(expand.grid(
  years_to_cfrd = -10:10,
  a1c_ogtt = unique(mod_df$a1c_ogtt),
  Age_YrEnd = mean(mod_df$Age_YrEnd, na.rm = T),
  Gender = "M",
  MutClass = "1-3",
  A_Vx770 = "No",
  A_VX809comb = "No"
))
pop_dat$x <- (pop_dat$years_to_cfrd >= 0) * pop_dat$years_to_cfrd
pred <- predict(m, re.form = NA, newdata = pop_dat, se.fit = T)
pop_dat$pred <- pred$fit
pop_dat$ll <- pop_dat$pred - (1.96 * pred$se.fit)
pop_dat$ul <- pop_dat$pred + (1.96 * pred$se.fit)
ggplot(
  pop_dat,
  aes(x = years_to_cfrd, y = pred, group = a1c_ogtt, colour = a1c_ogtt)
) +
  geom_point() +
  geom_errorbar(aes(ymin = ll, ymax = ul), width = 0.1) +
  geom_line() +
  xlab("Years from CFRD") +
  ylab("FVC % Predicted") +
  theme_classic(base_size = 15)
# Model table
m |> tbl_regression()
# Contrasts
neither_pre <- rep(0, nrow(summary(m)$coefficients))
neither_change <- rep(0, nrow(summary(m)$coefficients))
neither_pre[which(rownames(summary(m)$coefficients) == "years_to_cfrd")] = 1
neither_change[which(rownames(summary(m)$coefficients) == "x")] = 1
a1c_pre <- rep(0, nrow(summary(m)$coefficients))
a1c_change <- rep(0, nrow(summary(m)$coefficients))
a1c_pre[which(
  rownames(summary(m)$coefficients) %in%
    c("years_to_cfrd", "years_to_cfrd:a1c_ogttHbA1c Only")
)] = 1
a1c_change[which(
  rownames(summary(m)$coefficients) %in%
    c("x", "a1c_ogttHbA1c Only:x")
)] = 1
ogtt_pre <- rep(0, nrow(summary(m)$coefficients))
ogtt_change <- rep(0, nrow(summary(m)$coefficients))
ogtt_pre[which(
  rownames(summary(m)$coefficients) %in%
    c("years_to_cfrd", "years_to_cfrd:a1c_ogttOGTT")
)] = 1
ogtt_change[which(
  rownames(summary(m)$coefficients) %in%
    c("x", "a1c_ogttOGTT:x")
)] = 1
contr <- rbind(
  "HbA1c Only Pre-CFRD" = a1c_pre,
  "HbA1c Only Post-CFRD" = a1c_pre + a1c_change,
  "OGTT Pre-CFRD" = ogtt_pre,
  "OGTT Post-CFRD" = ogtt_pre + ogtt_change,
  "Neither Pre-CFRD" = neither_pre,
  "Neither Post-CFRD" = neither_pre + neither_change
)
change_mod_slopes <- glht(m, linfct = contr)
change_mod_slopes_ci <- confint(change_mod_slopes, calpha = univariate_calpha())
change_mod_slopes_ci$confint %>%
  as.data.frame() %>%
  gt(caption = "Slopes by Diagnosis Method", rownames_to_stub = T) %>%
  fmt_number(decimals = 2)
```
